Working with daily climate model output data in R and the \pkg{futureheatwaves} package

### 2016-12-29 paper_drafts/RJournalRevisions/anderson-eason-barnes.Rmd
Update article draft

## Customizing dates for projections and reference temperatures

By default, the `gen_hw_set` function will generate heat wave projections for the years 2070 to 2079, to align with the example data. This projection date range can be changed by specifying alternative starting and ending years in the `projectionBoundaries` argument of the `gen_hw_set` function. For example, to create projections for 2060 to 2079, the user would call:

```{r, eval = FALSE}
gen_hw_set(out = "example_results",
           dataFolder = projection_dir_location ,
           dataDirectories = list("historical" = c(1990, 1999),
                                        "rcp85" = c(2060, 2079)),
           citycsv = city_file_location,
           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
           tasFilenames = "tas_NorthAmerica_12mo.csv",
           timeFilenames = "time_NorthAmerica_12mo.csv",
           projectionBoundaries = c(2060, 2079))
```

The heat wave data sets characterize heat wave in several ways that are based on relative temperature (for example, the `mean.temp.quantile`, `max.temp.quantile`, and `min.temp.quantile` columns described in Table \ref{tab:hwcharacteristics}). These characteristics are measured for each of the heat waves identified by the `gen_hw_set` function by taking the absolute temperature of the heat wave (e.g., average temperature during the heat wave is 90\textsuperscript{o}F) and comparing it to the heat wave location's typical temperature distribution. This process generates relative measures of how intense the heat wave is compared to what is normal in that location (e.g., 90\textsuperscript{o}F is in the 99\textsuperscript{th} percentile of year-round temperatures in the location).

The values of these relative heat wave characteristics will depend on what time period is used to determine the location's typical temperature distribution. For example, the same heat wave would likely have much higher values for these relative characteristics if the temperature during the heat wave is compared to typical temperatures in the present day for that location versus if the temperature is compared to typical temperatures much later in the 21\textsuperscript{st} century for the same location.

The default in the `gen_hw_set` function is to use temperatures for the period 2070 to 2079 for these reference temperatures to calculate all relative heat wave characteristics. However, the user can change this specification using the `referenceBoundaries` option of `gen_hw_set`. This functionality can be useful in exploring the role of adaptation in future heat waves. For example, to use temperature projections for 1990 to 1999 when calculating relative characteristics of heat waves, to explore the assumption that cities remain adapted to their present-day climate, rather than changing in adaptation as climate change increases temperatures, a user could run:

```{r, eval = FALSE}
gen_hw_set(out = "example_results",
           dataFolder = projection_dir_location ,
           dataDirectories = list("historical" = c(1990, 1999),
                                        "rcp85" = c(2060, 2079)),
           citycsv = city_file_location,
           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
           tasFilenames = "tas_NorthAmerica_12mo.csv",
           timeFilenames = "time_NorthAmerica_12mo.csv",
           referenceBoundaries = c(1990, 1999))
```

For these date range specifications, there are some restrictions on which year ranges can be selected. The starting year cannot be earlier than the first year in the first subdirectory, the ending year cannot be later than the last year of the second subdirectory, and the custom date boundaries cannot span the two subdirectories.

@@

The functionality of this package can be easily expanded by loops. For example, to explore the role of the heat wave definition on projections, the user could create a loop to run `gen_hw_set` and `apply_all_models` to the same directory of climate projections but with a variety of different functions used to identify the heat waves in the projections.

Also, while this package was created to be used for research on heat waves in climate change projections, with some modifications it can be used more broadly. For example, there are other episodes like wildfires and air pollution where it may be interesting to identify extended periods of high exposures in projection time series, and this package could be applied to explore these exposures. The directory of projection data would need to be set up in the same structure as for exploring heat waves, and the `input_metric` should be set as `fahrenheit`, to pass the exposure values through to the characterized data sets without performing a conversion. A user could also use this package to explore events that must be lower than some minimum threshold (e.g., cold waves), but it would take some extra coding and conversions, since the functions in this package are written to identify periods above a threshold (for example, the user could multiple all projected temperatures by -1, and then the coldest temperatures would register as being the highest).

```{r eval = FALSE}
## Cold wave example
gen_hw_set(out = "example_results",
           dataFolder = projection_dir_location ,
           dataDirectories = list("historical" = c(1990, 1999),
                                  "rcp85" = c(2060, 2079)),
           citycsv = city_file_location,
           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
           tasFilenames = "tas_NorthAmerica_12mo.csv",
           timeFilenames = "time_NorthAmerica_12mo.csv", 
           probThreshold = 0.10, 
           above_threshold = FALSE, 
           absolute_thresholds = c(266, 263, 260, 258), 
           seasonal_months = c(12, 1, 2))
```

>  Extreme statistics and spell-lengths: "A 'frost day' is defined as one during which the minimum temperature falls below freezing point (0 degC). This is described as a climatological statistic, in which the minimum temperature is first calculated within each day, and then the number of days or spell lengths meeting the specified condition are evaluated." (from one example in http://cfconventions.org/Data/cf-conventions/cf-conventions-1.6/build/cf-conventions.html#calendar)

Users doing these kinds of extensions will need to pay attention to a few points. First, some of the event characteristics (first in the calendar year, average of May--September temperatures, days above 90\textsuperscript{o}F) might not be meaningful for studies of other types of events. Further, because event periods are usually defined as a string of multiple days exceeding some threshold, the functions in this package may miss the first and last event of the time period. For example, if the first day of the time series is the last day of an event, this function would not identify that event because it lacks data from the earlier days that allow this day to meet the event definition. This issue would lead to, at most, missing two events out of each projection, but should be considered if studying events that might occur near the start and end of projection data. If there is adequate interest from researchers, in the future we may adapt the package to make these secondary applications part of the package.

@@

This work was supported by grants from the National Institute of Environmental Health Sciences (R00ES022631), the National Science Foundation (1331399), and the Colorado State University Vice President for Research.

>>>>

# Customizing dates for projections and reference temperatures

Depending on the climate model output data used and the research question, different projection periods may be of interest. For example, one research may want to explore extreme events in the 2050--2075 period, while another may be interested in 2090--2100. The projection date range can be changed by specifying starting and ending years in the \code{projectionBoundaries} argument of the \code{gen\_hw\_set} function.

Further, some of the event characteristics (e.g., \code{mean.temp.quantile}, Table \ref{tab:hwcharacteristics}) are calculated based on relative temperature, which is a measure of how the value of the variable of interest during an extreme event compares to the typical distribution of that variable at that location. These characteristics are measured for each of the extreme events identified by the \code{gen\_hw\_set} function by taking the absolute value of the variable during the event (e.g., average temperature during the heat wave is 90\textsuperscript{o}F \ 32.2\textsuperscript{o}C) and comparing it to the location's typical variable distribution. This process generates relative measures of how intense the event is compared to what is normal in that location (e.g., 90\textsuperscript{o}F \ 32.2\textsuperscript{o}C is in the 99\textsuperscript{th} percentile of year-round temperatures in the location).

These relative event characteristics will vary depending on whether you calculate them based on a location's present-day variable distribution or on the location's variable distribution in the future, since the distributions of many relevant variables (e.g., temperature, precipitation) are expected to change in many locations with climate change. The package therefore allows the user to specify date ranges of the temperature distributions to be used in calculating these relative temperature metrics in each location, which can be done using the \code{referenceBoundaries} option of \code{gen\_hw\_set}.

@@

While this package was created to be used for research on extreme events in climate change projections, it can be used more broadly. For example, there are other episodes like wildfires and air pollution where it may be interesting to identify extended periods of high exposures in projection time series, and this package could be applied to gridded air pollution model output to explore these exposures.

@@

This work was supported by grants from the National Institute of Environmental Health Sciences (R00ES022631), the National Science Foundation (1331399), and the Colorado State University Vice President for Research. Thank you to early testers of the package:  Julia Bromberek, Wande Benka-Coker, Josh Ferreri, Ryan Gan, Molly Gutilla, Mike Lyons, Casey Quinn, Rachel Severson, and Meilin Yan.

### 2016-12-29 paper_drafts/RJournalRevisions/anderson-eason-barnes.tex
Update article draft

\subsection{Customizing dates for projections and reference
temperatures}\label{customizing-dates-for-projections-and-reference-temperatures}

By default, the \texttt{gen\_hw\_set} function will generate heat wave
projections for the years 2070 to 2079, to align with the example data.
This projection date range can be changed by specifying alternative
starting and ending years in the \texttt{projectionBoundaries} argument
of the \texttt{gen\_hw\_set} function. For example, to create
projections for 2060 to 2079, the user would call:

\begin{Schunk}
\begin{Sinput}
gen_hw_set(out = "example_results",
           dataFolder = projection_dir_location ,
           dataDirectories = list("historical" = c(1990, 1999),
                                        "rcp85" = c(2060, 2079)),
           citycsv = city_file_location,
           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
           tasFilenames = "tas_NorthAmerica_12mo.csv",
           timeFilenames = "time_NorthAmerica_12mo.csv",
           projectionBoundaries = c(2060, 2079))
\end{Sinput}
\end{Schunk}

The heat wave data sets characterize heat wave in several ways that are
based on relative temperature (for example, the
\texttt{mean.temp.quantile}, \texttt{max.temp.quantile}, and
\texttt{min.temp.quantile} columns described in Table
\ref{tab:hwcharacteristics}). These characteristics are measured for
each of the heat waves identified by the \texttt{gen\_hw\_set} function
by taking the absolute temperature of the heat wave (e.g., average
temperature during the heat wave is 90\textsuperscript{o}F) and
comparing it to the heat wave location's typical temperature
distribution. This process generates relative measures of how intense
the heat wave is compared to what is normal in that location (e.g.,
90\textsuperscript{o}F is in the 99\textsuperscript{th} percentile of
year-round temperatures in the location).

The values of these relative heat wave characteristics will depend on
what time period is used to determine the location's typical temperature
distribution. For example, the same heat wave would likely have much
higher values for these relative characteristics if the temperature
during the heat wave is compared to typical temperatures in the present
day for that location versus if the temperature is compared to typical
temperatures much later in the 21\textsuperscript{st} century for the
same location.

The default in the \texttt{gen\_hw\_set} function is to use temperatures
for the period 2070 to 2079 for these reference temperatures to
calculate all relative heat wave characteristics. However, the user can
change this specification using the \texttt{referenceBoundaries} option
of \texttt{gen\_hw\_set}. This functionality can be useful in exploring
the role of adaptation in future heat waves. For example, to use
temperature projections for 1990 to 1999 when calculating relative
characteristics of heat waves, to explore the assumption that cities
remain adapted to their present-day climate, rather than changing in
adaptation as climate change increases temperatures, a user could run:

\begin{Schunk}
\begin{Sinput}
gen_hw_set(out = "example_results",
           dataFolder = projection_dir_location ,
           dataDirectories = list("historical" = c(1990, 1999),
                                        "rcp85" = c(2060, 2079)),
           citycsv = city_file_location,
           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
           tasFilenames = "tas_NorthAmerica_12mo.csv",
           timeFilenames = "time_NorthAmerica_12mo.csv",
           referenceBoundaries = c(1990, 1999))
\end{Sinput}
\end{Schunk}

For these date range specifications, there are some restrictions on
which year ranges can be selected. The starting year cannot be earlier
than the first year in the first subdirectory, the ending year cannot be
later than the last year of the second subdirectory, and the custom date
boundaries cannot span the two subdirectories.

@@

The functionality of this package can be easily expanded by loops. For
example, to explore the role of the heat wave definition on projections,
the user could create a loop to run \texttt{gen\_hw\_set} and
\texttt{apply\_all\_models} to the same directory of climate projections
but with a variety of different functions used to identify the heat
waves in the projections.

Also, while this package was created to be used for research on heat
waves in climate change projections, with some modifications it can be
used more broadly. For example, there are other episodes like wildfires
and air pollution where it may be interesting to identify extended
periods of high exposures in projection time series, and this package
could be applied to explore these exposures. The directory of projection
data would need to be set up in the same structure as for exploring heat
waves, and the \texttt{input\_metric} should be set as
\texttt{fahrenheit}, to pass the exposure values through to the
characterized data sets without performing a conversion. A user could
also use this package to explore events that must be lower than some
minimum threshold (e.g., cold waves), but it would take some extra
coding and conversions, since the functions in this package are written
to identify periods above a threshold (for example, the user could
multiple all projected temperatures by -1, and then the coldest
temperatures would register as being the highest).

\begin{Schunk}
\begin{Sinput}
## Cold wave example
gen_hw_set(out = "example_results",
           dataFolder = projection_dir_location ,
           dataDirectories = list("historical" = c(1990, 1999),
                                  "rcp85" = c(2060, 2079)),
           citycsv = city_file_location,
           coordinateFilenames = "latitude_longitude_NorthAmerica_12mo.csv",
           tasFilenames = "tas_NorthAmerica_12mo.csv",
           timeFilenames = "time_NorthAmerica_12mo.csv", 
           probThreshold = 0.10, 
           above_threshold = FALSE, 
           absolute_thresholds = c(266, 263, 260, 258), 
           seasonal_months = c(12, 1, 2))
\end{Sinput}
\end{Schunk}

\begin{quote}
Extreme statistics and spell-lengths: ``A `frost day' is defined as one
during which the minimum temperature falls below freezing point (0
degC). This is described as a climatological statistic, in which the
minimum temperature is first calculated within each day, and then the
number of days or spell lengths meeting the specified condition are
evaluated.'' (from one example in
\url{http://cfconventions.org/Data/cf-conventions/cf-conventions-1.6/build/cf-conventions.html\#calendar})
\end{quote}

Users doing these kinds of extensions will need to pay attention to a
few points. First, some of the event characteristics (first in the
calendar year, average of May--September temperatures, days above
90\textsuperscript{o}F) might not be meaningful for studies of other
types of events. Further, because event periods are usually defined as a
string of multiple days exceeding some threshold, the functions in this
package may miss the first and last event of the time period. For
example, if the first day of the time series is the last day of an
event, this function would not identify that event because it lacks data
from the earlier days that allow this day to meet the event definition.
This issue would lead to, at most, missing two events out of each
projection, but should be considered if studying events that might occur
near the start and end of projection data. If there is adequate interest
from researchers, in the future we may adapt the package to make these
secondary applications part of the package.

>>>>

\section{Customizing dates for projections and reference
temperatures}\label{customizing-dates-for-projections-and-reference-temperatures}

Depending on the climate model output data used and the research
question, different projection periods may be of interest. For example,
one research may want to explore extreme events in the 2050--2075
period, while another may be interested in 2090--2100. The projection
date range can be changed by specifying starting and ending years in the
\code{projectionBoundaries} argument of the \code{gen\_hw\_set}
function.

Further, some of the event characteristics (e.g.,
\code{mean.temp.quantile}, Table \ref{tab:hwcharacteristics}) are
calculated based on relative temperature, which is a measure of how the
value of the variable of interest during an extreme event compares to
the typical distribution of that variable at that location. These
characteristics are measured for each of the extreme events identified
by the \code{gen\_hw\_set} function by taking the absolute value of the
variable during the event (e.g., average temperature during the heat
wave is 90\textsuperscript{o}F ~32.2\textsuperscript{o}C) and comparing
it to the location's typical variable distribution. This process
generates relative measures of how intense the event is compared to what
is normal in that location (e.g., 90\textsuperscript{o}F
~32.2\textsuperscript{o}C is in the 99\textsuperscript{th} percentile of
year-round temperatures in the location).

These relative event characteristics will vary depending on whether you
calculate them based on a location's present-day variable distribution
or on the location's variable distribution in the future, since the
distributions of many relevant variables (e.g., temperature,
precipitation) are expected to change in many locations with climate
change. The package therefore allows the user to specify date ranges of
the temperature distributions to be used in calculating these relative
temperature metrics in each location, which can be done using the
\code{referenceBoundaries} option of \code{gen\_hw\_set}.

@@

While this package was created to be used for research on extreme events
in climate change projections, it can be used more broadly. For example,
there are other episodes like wildfires and air pollution where it may
be interesting to identify extended periods of high exposures in
projection time series, and this package could be applied to gridded air
pollution model output to explore these exposures.
